Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-05 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Close to complete on RSE sections

Don Middleton -- National Center for Atmospheric Research
Joshua Greenberg -- Alfred P. Sloan Foundation
James Hetherington -- University College London
Lindsay Powers -- The HDF Group
Mark Miller
Dan Sellars: CANARIE

\begin{itemize}
\item Don Middleton -- National Center for Atmospheric Research
\item Josh Greenburg -- affiliation2
\end{itemize}

\subsection{Temporary direct paste-in of notes from Workshop}

Our charge: document successful funding models as a means of creating a conversation that can be used in finding funding for new efforts;
Don Middleton
Josh Greenburg: SLOAN
James Hetherington: University College London j.hetherington@ucl.ac.uk
Lindsay Powers: The HDF Group, lpowers@hdfgroup.org
Mark Miller
Dan Sellars: CANARIE: Canada; like internet 2, they make some funding available to do projects.
Fund a PI /team to produce a platform to do the work.
Cultural barrier is steep for communicating across barriers, and scientists seeing the developers as “other”. SO collaboration, and tools to fund/encourage/motivate collaboration are extremely important.
Projects that engage communities on an “at will” basis
One model is to engage scientists in the development process as part of a community effort, and that may be largely unfunded. Models that are community driven can be very organic, so new contributions are possible from unexpected places.
Funded engagement Models
It is usually more effective and efficient to do development where there is funding to support the collaboration.
Co-funding on a single proposal.
One way to do this is by co-funding individuals from both camps, that is domain scientists and developers are funded together in a single proposal. At NCAR, University partnerships are required for submission of proposals, so collaboration is an essential part of grant submission, and this will tend to bring developers and scientists together. This is certainly an effective method for promoting project based/project scale work. We aren’t sure how this requirement could be imposed globally, but it should and probably often is required by any given RFP for SW development. One example of a success is the development of semantics and linked data in support of Ocean Sciences. An Earthcube funded project pairs domain scientists with RSEs and has been successful as the semantics attached have increased data use and discovery significantly.
Developer support from a resource provider to augment existing funding.
We discussed another fundamental strategy used by several organizations, where a group that wants to promote use of a new resource, be it internet (Canarie) or one or more compute engines (XSEDE (US) and Archer (UK)) fund a developer team that works directly with scientists from outside the organization to promote the use of the new resource.

@@

The UK allows this kind of collaboration to support the creation of scientific software for the large supercomputing resource called Archer. The Archer project fund a group of developers, and domain scientists can apply to the group for developer time to help them launch a new tool on Archer.
The US XSEDE project does essentially the same thing. The extended collaboration service (ECS) is a set of developers who are paid with XSEDE funding, and are on “permanent” staff.  When PIs request allocations on XSEDE resources, there is a finite pool of developer time that can be awarded, typically for one year only, and at partial effort, typically 20% or so. The finite time allowed provides motivation for the scientist and their group to work closely the developer and to become educated in what the developer is doing, so they can sustain the effort, once the ECS period is over.
This funding mechanism is highly efficient for scientific problems, because the developer pool assembled by the research providers are, by definition, expert in the characteristics of their specific resource, and can very quickly assess the scientists needs, and what it will take to implement software that meets the user’s needs. Moreover, when an organization assembles a developer pool, there will be diversity in the talent pool, so issues that require multiple kinds of expertise can be addressed by distinct individuals, again with maximum efficiency.
Promoting institutional support for research science developers
The Sloan Foundation is currently funding data science engineers, who work in the context of other SW developers at University of Washington.  These scientists work in the e-Science Studio/dataScience Studio, and they help a group of graduate students in solving their problems in data science/data management. During Fall and Spring, a 10-week incubator program allows students to work two days a week to work on a data-intensive science project. Some fraction of the developer time is dedicated to the developers personal interests as well as instruction. Piloting and documenting what works and what doesn’t.

@@

Researchers wont be able to do this without seed money to get it going. James succeeded at his institution which was building cyberinsfrastructure. Three posts were funded by the university. This is budget line for professional developers. Currently developer support for individual projects is awarded by a panel. Researchers can contribute funds that are left over from other awards where people leave prematurely. This is happening currently.  Some people committed to using funds in an existing award. There is a UK fellowship for replicating this basic concept: the award is for salary for one position as team leader,  and a first team member; these awards are for five years duration (woohoo!). There are other RSEs in other UK institutions. These may engage in labor exchange via  informal work sharing agreements between institutions. Three month engagement preferred, but not everyone works that way some prefer quick in and done modes.
To make such a program work, it seems it  must be university wide, to be broad enough to have the ftes needed funded. But there is not one mind on this, it maybe that the broadest approach is not idea. Perhaps a specialization focus on developers to support physical or biological sciences may be preferable, if the customer base is large enough. How to aggregate enough work to make it sustainable. How do you make sure you have the right staff skills? This will be an issue in a smaller group, and it will be exacerbated if the approach is general rather than focused on a domain emphasis where the number of possible skills required is less. The optimum use of resources seems to be 40% effort, on a project; two projects per developer. Two developers per project seems to be ideal, in the sense that software development is enhanced by two pairs of eyes..
Collaborative computational projects (CCP) to support community networking: domain specific communities put forward proposals that are a priority of the community as a whole: biosimulation plasma PHYSICS QUANTUM PHYSICS. % years of funding. Custodians of community codes. Software support. help these are incubators. Add new features. Software of the future, open to all. Ccp submits. 15 ftes for development across all CCPs. GET IAIN. 4% success rate for software of the future project, high end computing consortia, in domains of CCPs. Co-design; the art of bringing it into existence together. The mission is codesigning 5 projects: the collaboration did work, but not as intimately as it might have.  Incentive structures are still based on getting papers out, and the benefits to the postdocs are career focused.
RSE fellowships: 204 applications, 5 funded. Intention was to fund applicants that demonstrated broad benefit to community/ies. How to better incentivize the collaboration to serve a broader community? Challenge: Scientists do not appreciate the engineering issues, whereas the cs person does not necessarily have the understanding of the domain.
Possibility of a training course for research software engineering, a course could be run to teach people to manage a grant that is focused on software development. Non-hands on software carpentry for PIs.
What are the next steps:
Have we actually been successful?
What are the metrics for success?
Publications are tracked in James organization.
How to evaluate the productivity gains that are made as a result of this engagement with James’ group?
How to evaluate what did not go wrong because of this intervention?
Notre Dame and LSU groups who are RSE groups.
NCEAS/SOCSYNC; some RSE support; NESCENT/NIMBIOS also.
The power of the label? Getting the word out about RSE support using the label.

Will research science developers be required, in the long run?
One issue that came up was whether the need for developers was a time bounded one; is it the case that the new generation of computer and software savvy scientists will be so comfortable in developing their own code that the professional developer will not be needed.  And this brings up the flip side question, “Do scientists really want to be writing code?”
Career Path.
We also had a little discussion about how to make a career path for research developers. It need not be solely an academic enterprise, but tenure is always problematic for people of this class. There is a need for more senior people to have a track. Retention is hard, because these people are highly in demand by the economy, so retention is hard. People who have a PhD in science and enter industry, may return for diverse reasons.
Things we haven’t started talking about:
Other above the campus models for funding;
Industry:academic partnerships

Follow-on Discussion at GCE15
rse.ac.uk
People: Lorraine Hwang, Smon Trigger, Nancy Wilkins-Diehr,  Alex Vyshkov, Sandra, Keynote person from Zooniverse

Good example: Notre Dame (14 research programmer), LSU, Princeton,
SDSC -- actually even in national labs there are inefficiencies with per-project jobs rather than develop pool -- potential issue is degree of control.
Matt Trunnel - CIO of Broad institute boston - Chris Dwan - they had individual RSEs, then aggregated.
James Taylor’s team  - John’s Hopkins
Org models - OVPR, CS, Physics, IT, Library
IT Coalescence issues.
Maybe Yale?

\subsection{Description of Opportunity, Challenges, and Obstacles}

@@

\subsection{Plan for Future Organization}

\subsection{What Else is Needed?}

\subsection{Key Milestones and Responsible Parties}

@@

>>>>

The group at WSSPE:

\begin{itemize}
\item Don Middleton -- National Center for Atmospheric Research
\item Joshua Greenberg -- Alfred P. Sloan Foundation
\item James Hetherington -- University College London
\item Lindsay Powers -- The HDF Group
\item Mark A. Miller -- San Diego Supercomputer Center
\item Dan Sellars -- CANARIE
\end{itemize}

This was further enhanced by additional discussions at the following
GCE15 conference:

\begin{itemize}
\item Lorraine Hwang -- UC Davis
\item Simon Trigger
\item Nancy Wilkins-Diehr -- San Diego Supercomputer Center
\item Alexander Vyushkov -- Notre Dame
\item Sandra Gesing - Notre Dame
\item Ali Swanson -- University of Oxford

\end{itemize}

In addition to the points noted in the main discussion~\ref{RSE}, we also
discussed the following:

``Are you an RSE or a RA?'' -- this is not properly a binary question. Most of
us sit at different points on that spectrum, and move along it during our
careers. (Ususally from RA to RSE -- examples of movement in the other direction
from readers would be welcomed.)
Either way, the label ``Research Software Engineer'' is now starting to
have some power. Many scientists do not want to be writing code; some do, to
varying degrees. These groups can usefully support each other.

he power of the label? Getting the word out about RSE support using the label.

Will research science developers be required, in the long run?
One issue that came up was whether the need for developers was a time bounded one; is it the case that the new generation of computer and software savvy scientists will be so comfortable in developing their own code that the professional developer will not be needed.  And this brings up the flip side question, “Do scientists really want to be writing code?”
Career Path.
We also had a little discussion about how to make a career path for research developers. It need not be solely an academic enterprise, but tenure is always problematic for people of this class.

Skills and resources may vary between teams. To help resolve this, maintaining
high levels of communication between groups will be valuable. In the UK, there
are plans to permit resource sharing between institutional RSE groups.

Collaborative funding can be important for RSE groups, to ensure that research
leadership remains with the domain scientists. At NCAR, University partnerships
are required for submission of proposals, so collaboration is an essential part
of grant submission, and this will tend to bring developers and
scientists together. The UCL group also follows this approach, with all bids
requiring an academic collaborator.
Domain scientists and developers are funded together in a single proposal.
Another example of a success is the development of semantics and linked data in
support of Ocean Sciences. An Earthcube funded project pairs domain scientists
with RSEs and has been successful as the semantics attached have increased data
use and discovery significantly.

An alternative approach has been the provision of programming expertise as
part of national compute services. The US XSEDE project's extended collaboration
 service (ECS) is a set of developers who are paid with XSEDE funding,
 and are on “permanent” staff.  When PIs request allocations on XSEDE resources,
 there is a finite pool of developer time that can be awarded,
typically for one year only, and at partial effort, typically 20 percent or so.
The finite time allowed provides motivation for the scientist and their group to
work closely the developer and to become educated in what the developer is doing,
so they can sustain the effort, once the ECS period is over. This funding
mechanism can be highly efficient for scientific problems, because the developer
pool assembled by the research providers are, by definition, expert in the
characteristics of their specific resource, and can very quickly assess the
scientists needs, and what it will take to implement software that meets the
user’s needs. However, it does not develop capacity within institutions,
and indeed can have a ``chilling effect''.

The UK allows this kind of collaboration to support the creation of scientific
software for the large supercomputing resource called ARCHER.
However, as well as resourcing this from the staff of the Edinburgh Parallel
Computing Centre, who host the computer, this ``embedded CSE'' resource also
allows the programming to come from local groups. This has been very helpful
in providing funding to establish local groups. These work best when they
develop good collaborations with national cyberinfrastucture pools.
When an organization assembles
a developer pool, diversity is developed and skills can be transferred.

We would like to see these models applied outside High Performance Computing;
most scientific software is not destined to run on national cyberinfrastructure,
but needs similar support. The argument regarding making better use of expensive
hardware through software improvements has been useful politically, (and many
RSE groups are cited in organisations which host clusters for this reason), but
the time has come to make the case that software itself is a critical
cyberinfrastructure, and, with a much longer shelf-life than hardware, is itself
a capital investment.

@@

There can be problems communicating across cultural barriers, with domain
scientists seeing developers as “other”.
Collaboration, and tools to fund, encourage or motivate collaboration are
extremely important.

We think support from non-governmental organisations will be important if RSE
groups will become established.
The Sloan Foundation is currently funding data science engineers, who work in
the context of other SW developers at University of Washington.
These scientists work in the e-Science Studio/dataScience Studio, and they
help a group of graduate students in solving their problems in data
science and data management. During Fall and Spring, a 10-week incubator program
allows students to work two days a week to work on a data-intensive science
project. Some fraction of the developer time is dedicated to the developers'
personal interests as well as instruction.

@@

One model that has been helpful in finding funding for RSE groups is the use of
funds left over on research grants when RAs have left prematurely -- PIs like
this arrangement as it is hard to find good staff for short-term positions, so
having a pool of research programming staff on hand resolves this problem.
We recommend that funders give explicit guidance to grant holders and
institutions that such arrangement are favourable. Framework agreements
permitting this to go ahead without
checking back every time with funders and/or grant panels would further
smooth this.
(This is also provides a more stable job for those who hold these skills, but
arguments about making life nicer for postdocs will not help persuade funders
or PIs!)

There is some question about the most effective
duration and percent of full time for a programmer's work on a project.
At least three months is necessary for the programmer to read into the science
(RSEs must not become so disengaged from research that they don't have time to
read a few papers -- this will result in code which doesn't meet scientific
needs), but too long could result in an RSE losing their flexibility, becoming
so engaged in one project that when that project ends, they find it hard to
transfer. For this reason, we also recommend that 40 percent is ideal;
two projects per developer, with some time for training and infrastructure work.
Two developers per project seems to be ideal, in the sense that software
development is enhanced by two pairs of eyes.

There is as yet no clear answer as to the scale of aggregation needed to
make such a program work. A university wide programme allows enough scale to
be robust to fluctuations of funding within one field.
But a specialization focus on developers to support, for example,
physical or biological
sciences may be preferable, if the customer base is large enough.
The desire to aggregate enough work to make it sustainable, and the need to
have domain-relevant research programming skils, are in tension.

In the UK, another source of funding for research software has been the
Collaborative computational projects (CCPs): domain specific communities put
forward proposals that are a priority of the community as a whole, for example,
biosimulation or plasma physics. These bodies act as custodians of community codes,
and a central team also provides software engineering support support.

However this area develops, the need for funding to secure software as a
cyberinfrastructure component is clear. Funding which permits code to be
refactored, tidied and optimised is rare; this is often done ``on the sly''
in a scientifically focused grant. The UK EPSRC's ``software for the future''
call, which really permits explicit investiment in software as an infrastructure,
is so oversubscribed as to have a FOUR PERCENT success rate; the demand is clear!

One opportunity is the idea of co-design; where infrastructural libraries are
developed alongside the scientific codes that will call them. However,
collaboration is hard to foster here; as incentive structures are still focused
on short-term papers. This can cause infrastructure developers to
focus more on publications
in their areas of mathematics and CS, the domain developers on the shorter-term
needs of their own fields. Genuine collaborative co-construction
is harder to foster.

It can be a more difficult
to help leading domain scientists see the value of engineering effort than those
in their teams who are forced to work with difficult-to-use or unreliable
software tools, as they do not see the pain. Perhaps a version of
``software carpentry'' targeted at those PIs
who are awarded or apply for software heavy grants could be valuable here.

RSEs provide a useful contribution to their Universities' *teaching* missions,
as well as for research, as they are well placed to deliver the research programming
training that many scientists now need. In the
longer term, with programming skills taught to all through their career,
we hope specialist
scientific developers will be less needed.

@@

We will seek to identify and approach existing research programming organisations,
to get their permission to list them on a list of research software groups.
We will also look for examples of groups which have successfully become self-
sustaining following

\subsection{Plan for Future Organization and Future Needs}

The UKRSE community will provide initial facilities to host this list, and
continue to work to spread the initiative, but local leadership in the US
is needed if this campaign is to succeed. This will require an initial gathering
of identified research software organisations in the US to this end.

@@

Financial support for an initial conference to bring together research software
groups to form an organisation and create a resource sharing structure would
help to further this campaign.

In the longer term, funding organisations, especially non-governmental organisations
with the capability to effect innovation through seed funding, could provide
support to nucleate the creation of research software groups.

### 2015-11-05 WSSSPE3/report/wg_main_funding_spec_expert.tex
Close to complete on RSE sections

\subsection{Research Programming Expertise -- how to fund it?}

@@

One solution to this problem would be a successful solution to the problems of
obtaining academic credit for programming work. If this were solved, traditional
academic funding models and career tracks would properly sustain this
contribution to research. However, in the absence of guaranteed success in that
stream, this working party seeks to look to an alternate solution; the
development and funding of dedicated research programming roles. We also noted
that measures of impact and success for research programming
groups, as well as for individual research software engineers, will be required
in order to make the case for continued funding.

A contributing factor to the problem is the devastating effect on an academic
publication record resulting from time in industry; such postings often
develop exactly the skills that research programmers need, yet returns to
university positions following an industry role are penalised by the current
structures.

@@

One solution is to fund dedicated programming roles for major research software
projects at national laboratories or other non-educational institutions. This
solution is working well for many well-used scientific codebases.
However, much of the body of software for which this expertise is needed sits
within the longer tail of research in universities. Therefore, we argue that
institutions need to develop hybrid academic-technical roles for this capacity
who work with more than one PI, rather than the traditional RA role within a
single group. This could be centrally, perhaps within research computing
organisations which have traditionally supported university cyberinfrastructure,
or more locally.

@@

@@

TBD, probably into the UKRSE site.

>>>>

\subsection{Research Programming Expertise -- how to fund it?}
\label{RSE}

@@

One solution to this problem would be a successful solution to the problems of
obtaining academic credit for programming work. If this were solved, traditional
academic funding models and career tracks would properly sustain this
contribution to research. However, in the absence of guaranteed success in that
stream, this working party seeks to look to an alternate solution; the
development and funding of dedicated research programming roles. We also noted
that measures of impact and success for research programming
groups, as well as for individual research software engineers, will be required
in order to make the case for continued funding. Research programmers will not
be measured by publications, we hope, but by other measures. Middle-author
publications are common for RSEs. Most RSEs welcome co-authorship on papers
where the PIs consider the contribution deserves it.

A contributing factor to the problem is the devastating effect on an academic
publication record resulting from time in industry; such postings often
develop exactly the skills that research programmers need, yet returns to
university positions following an industry role are penalised by the current
structures. Retention of senior programmers is hard, because these people
are highly in demand by the economy. However,
people who have a PhD in science and enter industry,
may desire to return for diverse reasons, and should be welcomed back.

@@

One solution is to fund dedicated programming roles for major research software
projects at national laboratories or other non-educational institutions. This
solution is working well for many well-used scientific codebases.
However, much of the body of software for which this expertise is needed sits
within the longer tail of research in universities. Therefore, we argue that
institutions need to develop hybrid academic-technical roles for this capacity
who work with more than one PI, rather than the traditional RA role within a
single group. This could be centrally, perhaps within research computing
organisations which have traditionally supported university cyberinfrastructure,
library organisations, or research offices. Alternatively, these groups could
be organisationally closer to research groups, sitting within academic
departments. The most
effective model will vary from institution to institution,
but the mandate and ways of working should be similar.

@@

Success, appropriately measured, will help make the case for further investment.
However, measurement of code cleanliness, and the resulting productivity gains,
is an unsolved problem in empirical software engineering. To measure
``what did not go wrong'' because of an intervention is particularly hard.

@@

http://www.rse.ac.uk/groups
